---
title: "Real-World Machine Learning"
subtitle: "Chapter 5"
author: "Paul Adamson"
date: "April 2, 2017"
output: html_document
---

This notebook contains R code to accompany Chapter 5 of the book 
["Real-World Machine Learning"](https://www.manning.com/books/real-world-machine-learning),
by  Henrik Brink, Joseph W. Richards, and Mark Fetherolf.  The code was 
contributed by [Paul Adamson](http://github.com/padamson). 

*NOTE: working directory should be set to this file's location.*

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(bit64)
library(dplyr)
library(plyr)
library(caret)
library(randomForest)
library(rhandsontable)
```

### Data for event-recommendation model

The data for the examples from this chapter are available at the
[Kaggle Event Recommendation Engine Challenge](https://www.kaggle.com/c/event-recommendation-engine-challenge).
The rules of the challenge prohibit redristribution of the data. 
To run the example code, login to [Kaggle](https://www.kaggle.com), 
download the data, and put the `train.csv`, `events.csv`, and `users.csv` 
files in a folder named `../proprietary_data/`.

### Figure 5.2 A sample of the datasets used for training the event-recommendations model

```{r loadEventData, cache=TRUE}
dataDir <- "../proprietary_data"
# join on user and event identifiers
#from users data, get user_id, birthyear, gender, timezone
users <- read.csv(file.path(dataDir, "users.csv"),
                  colClasses = c('integer64', 
                                 "NULL", 
                                 "integer",
                                 "factor",
                                 "NULL",
                                 "NULL",
                                 "integer"),
                  na.strings = c("16-Mar", "23-May", "NA", "None"))

# from events data, get event_id, lat, lng
# note don't get confused about the user_id in this dataset. it is the user that created the event and is different from the user_id in the train dataset
events <- fread(file.path(dataDir, "events.csv"),
                sep=",",
                colClasses = c("integer64",
                               rep("NULL",6),
                               "numeric",
                               "numeric",
                               rep("NULL",101)))

# from train data, get user, event, invited, interested, not_interested 
# rename user to user_id and event to event_id to match users and events datasets
# also need to drop observations without explicit interested or not_interested selection
# then drop not_interested since we only need one of the categorical variables
train <- read.csv(file.path(dataDir, "train.csv"),
                  colClasses = c("integer64",
                                 "integer64",
                                 "integer",
                                 "NULL",
                                 "integer",
                                 "integer"),
                  col.names = c("user_id",
                                "event_id",
                                "invited",
                                "NULL",
                                "interested",
                                "not_interested")) %>%
  subset(interested == 1 | not_interested == 1) %>%
  select(-not_interested)
```

```{r figure5.2, cache=TRUE, dependson="loadEventData"}
train.joined <- inner_join(train, users, by=c("user_id")) %>%
  inner_join(events, by=c("event_id")) %>%
  select(-c(user_id,event_id))
kable(head(train.joined,7))
```

## Figure 5.3 Cross-validated ROC curve and AUC metric for the simple event- recommendation model
```{r figure5.3_dummyvars, cache=TRUE, dependson="figure5.2"}
dummies <- dummyVars(" ~ .", data = train.joined, fullRank = TRUE, 
                     levelsOnly = TRUE)
train.joined.numeric <- data.frame(predict(dummies, newdata = train.joined)) %>%
  select(c(-female)) %>%
  na.omit()
train.joined.numeric$interested <- factor(train.joined.numeric$interested)
train.joined.numeric$interested <- 
  revalue(train.joined.numeric$interested, c("0" = "no", "1" = "yes"))
```

```{r chapter4listings, echo=FALSE}
read_chunk('chapter4listings.R')
```

```{r listing4.3, echo=TRUE,tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<listing4.3>>
```

```{r listing4.4, echo=TRUE,tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<listing4.4>>
```

```{r plotROC, eval=TRUE, cache=TRUE}
plotROC <- function(trueLabels, predictedProbs, nPoints=100, posClass=1){
  auc <- auc(trueLabels, predictedProbs, nPoints, posClass)
  df <- rocCurve(trueLabels = trueLabels,
                 predictedProbs = predictedProbs,
                 nPoints = nPoints,
                 posClass = posClass)
  ggplot(df,aes(x=fpr,y=tpr)) +
    geom_step(direction="vh") +
    scale_x_continuous(limits = c(0,1)) +
    scale_y_continuous(limits = c(0,1)) +
    labs(x = "False-positive rate",
         y = "True-positive rate") +
    annotate("text", x=.5,y=.5,label=paste0("Area under the curve: ",round(auc,digits = 2)))
  
}
```


```{r figure5.3_data, cache=TRUE, dependson="figure5.3_dummyvars"}
trainIndex <- createDataPartition(train.joined.numeric$male, p = .8, 
                                  list = FALSE, 
                                  times = 1)

eventTrain <- train.joined.numeric[ trainIndex,]
eventTest  <- train.joined.numeric[-trainIndex,]


fitControl <- trainControl(method = "repeatedcv", 
                           number = 10, 
                           savePredictions = TRUE,
                           summaryFunction=twoClassSummary, 
                           classProbs=TRUE)

rfFit <- train(interested ~ ., data = eventTrain,
                method = "rf", 
                trControl = fitControl, 
                verbose = FALSE)

event.rf.pred.prob <- predict(object=rfFit, 
                              eventTest, type='prob')
```

```{r figure5.3, cache=TRUE, dependson=c("figure5.3_data","plotROC")}
plotROC(revalue(eventTest$interested, c("no" = 0, "yes" = 1)),
        event.rf.pred.prob$yes, nPoints = 1000)
```

##  Figure 5.4 Additional date-time columns extracted from the timestamp column for the event-recommendation dataset
```{r figure5.4, cache=TRUE, dependson="loadEventData"}
# from events data, get event_id, start_time, lat, lng
# note don't get confused about the user_id in this dataset. it is the user that created the event and is different from the user_id in the train dataset
# also, doing a semi_join so we only work with the event data in the train data
# frame

events <- fread(file.path(dataDir, "events.csv"),
                sep=",",
                colClasses = c("integer64",
                               "NULL",
                               'myDateTime',
                               rep("NULL",4),
                               "numeric",
                               "numeric",
                               rep("NULL",101))) %>%
  semi_join(y = train) %>%
  mutate(datetime = strptime(start_time, format="%Y-%m-%dT%H:%M:%OS", tz="UTC"),
         datetime_year = as.numeric(substr(datetime, 1, 4)), 
         datetime_month_of_year = as.numeric(substr(datetime, 6, 7)), 
         datetime_day_of_month = as.numeric(substr(datetime, 9,10)), 
         datetime_hour_of_day = as.numeric(substr(datetime, 12, 13)), 
         datetime_minute_of_hour = as.numeric(substr(datetime, 15, 16)),
         datetime_second_of_minute = as.numeric(substr(datetime, 18, 19)),
         datetime_day_of_week = as.factor(weekdays(datetime)),
         datetime_week_of_year = week(datetime),
         datetime_quarter_of_year = quarter(datetime),
         datetime_day_of_year = as.POSIXlt(datetime)$yday
  ) %>%
  select(c(-datetime, -start_time))

kable(head(
  select(events,
         c(datetime_hour_of_day, 
           datetime_day_of_week,
           datetime_day_of_month,
           datetime_day_of_year,
           datetime_month_of_year)),
  5))

kable(head(
  select(events,
         c(datetime_minute_of_hour,
           datetime_second_of_minute,
           datetime_year,
           datetime_quarter_of_year,
           datetime_week_of_year)), 
  5))
```

## Figure 5.5 Cross-validated ROC curve for model including date-time features

```{r figure5.5_data, cache=TRUE, dependson=c("figure5.4","plotROC")}
train.joined <- inner_join(train, users, by=c("user_id")) %>%
  inner_join(events, by=c("event_id")) %>%
  select(-c(user_id,event_id))

dummies <- dummyVars(" ~ .", data = train.joined, fullRank = TRUE, 
                     levelsOnly = TRUE)
train.joined.numeric <- data.frame(predict(dummies, newdata = train.joined)) %>%
  select(c(-female)) %>%
  na.omit()
train.joined.numeric$interested <- factor(train.joined.numeric$interested)
train.joined.numeric$interested <- 
  revalue(train.joined.numeric$interested, c("0" = "no", "1" = "yes"))

trainIndex <- createDataPartition(train.joined.numeric$male, p = .8, 
                                  list = FALSE, 
                                  times = 1)

eventTrain <- train.joined.numeric[ trainIndex,]
eventTest  <- train.joined.numeric[-trainIndex,]


fitControl <- trainControl(method = "repeatedcv", 
                           number = 10, 
                           savePredictions = TRUE,
                           summaryFunction=twoClassSummary, 
                           classProbs=TRUE)

rfFit <- train(interested ~ ., data = eventTrain,
                method = "rf", 
                trControl = fitControl, 
                verbose = FALSE)

event.rf.pred.prob <- predict(object=rfFit, 
                              eventTest, type='prob')
```

```{r figure5.5, cache=TRUE, dependson=c("figure5.5_data","plotROC")}
plotROC(revalue(eventTest$interested, c("no" = 0, "yes" = 1)),
        event.rf.pred.prob$yes, nPoints = 1000)
```


##  Figure 5.6 A slice of the bag-of-words data for the event-recommendation example. These numbers are the counts of the top-occurring words in the event descriptions. A large fraction of the cells contain 0, so we call the dataset sparse.
```{r figure5.6_data, cache=TRUE, dependson="loadEventData"}
# from events data, get event_id, start_time, lat, lng, c_1 through c_other
# note don't get confused about the user_id in this dataset. it is the user that created the event and is different from the user_id in the train dataset
# also, doing a semi_join so we only work with the event data in the train data
# frame

events <- fread(file.path(dataDir, "events.csv"),
                sep=",",
                colClasses = c("integer64",
                               "NULL",
                               'myDateTime',
                               rep("NULL",4),
                               "numeric",
                               "numeric",
                               rep("integer",101))) %>%
  semi_join(y = train) %>%
  mutate(datetime = strptime(start_time, format="%Y-%m-%dT%H:%M:%OS", tz="UTC"),
         datetime_year = as.numeric(substr(datetime, 1, 4)), 
         datetime_month_of_year = as.numeric(substr(datetime, 6, 7)), 
         datetime_day_of_month = as.numeric(substr(datetime, 9,10)), 
         datetime_hour_of_day = as.numeric(substr(datetime, 12, 13)), 
         datetime_minute_of_hour = as.numeric(substr(datetime, 15, 16)),
         datetime_second_of_minute = as.numeric(substr(datetime, 18, 19)),
         datetime_day_of_week = as.factor(weekdays(datetime)),
         datetime_week_of_year = week(datetime),
         datetime_quarter_of_year = quarter(datetime),
         datetime_day_of_year = as.POSIXlt(datetime)$yday
  ) %>%
  select(c(-datetime, -start_time))
```

```{r figure5.6, cache=TRUE, dependson="figure5.6_data"}
rhandsontable(matrix(data=unlist(events[1:110,4:104]),nrow=11,ncol=10),
              readOnly = TRUE)
```

## Figure 5.7 Cross-validated ROC curve for full model including date-time and text features
```{r figure5.7_data, cache=TRUE, dependson="figure5.6_data"}
train.joined <- inner_join(train, users, by=c("user_id")) %>%
  inner_join(events, by=c("event_id")) %>%
  select(-c(user_id,event_id))

dummies <- dummyVars(" ~ .", data = train.joined, fullRank = TRUE, 
                     levelsOnly = TRUE)
train.joined.numeric <- data.frame(predict(dummies, newdata = train.joined)) %>%
  select(c(-female)) %>%
  na.omit()
train.joined.numeric$interested <- factor(train.joined.numeric$interested)
train.joined.numeric$interested <- 
  revalue(train.joined.numeric$interested, c("0" = "no", "1" = "yes"))

trainIndex <- createDataPartition(train.joined.numeric$male, p = .8, 
                                  list = FALSE, 
                                  times = 1)

eventTrain <- train.joined.numeric[ trainIndex,]
eventTest  <- train.joined.numeric[-trainIndex,]


fitControl <- trainControl(method = "repeatedcv", 
                           number = 10, 
                           savePredictions = TRUE,
                           summaryFunction=twoClassSummary, 
                           classProbs=TRUE)

rfFit <- train(interested ~ ., data = eventTrain,
                method = "rf", 
                trControl = fitControl, 
                verbose = FALSE)

event.rf.pred.prob <- predict(object=rfFit, 
                              eventTest, type='prob')
```

```{r figure5.7, cache=TRUE, dependson=c("figure5.7_data","plotROC")}
plotROC(revalue(eventTest$interested, c("no" = 0, "yes" = 1)),
        event.rf.pred.prob$yes, nPoints = 1000)
```

## Inset: Some algorithms have built-in feature selection

```{r rf.importance, cache=TRUE, dependson="figure5.7_data"}
rfImp <- varImp(rfFit, scale = FALSE)
setDT(rfImp$importance, keep.rownames = TRUE)[]
rfImp$importance <- 
  rfImp$importance[order(rfImp$importance$Overall,
                         decreasing = TRUE),]
rfImp$importance$rn <- reorder(rfImp$importance$rn, rfImp$importance$Overall)
ggplot(rfImp$importance[1:7], aes(x = rn, y = Overall)) + 
  geom_bar(stat = "identity") +
  xlab("Feature") +
  ylab("Importance") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  theme(text=element_text(size=16, family="Arial")) +
  scale_y_continuous(position="top") +
  #theme(axis.title.y = element_text(angle = 90, vjust = 1.08)) +
  coord_flip()
```