---
title: "Real-World Machine Learning"
subtitle: "Chapter 7"
author: "Paul Adamson"
date: "May 20, 2017"
output: html_document
---

This notebook contains R code to accompany Chapter 7 of the book 
["Real-World Machine Learning"](https://www.manning.com/books/real-world-machine-learning),
by  Henrik Brink, Joseph W. Richards, and Mark Fetherolf.  The code was 
contributed by [Paul Adamson](http://github.com/padamson). 

*NOTE: working directory should be set to this file's location.*

```{r setup, include=FALSE}
set.seed(1111)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(RSocrata)
library(xts)
#library(data.table)
library(plyr)
library(dplyr)
library(lubridate)
#library(ggplot2)
#library(gridExtra)
#library(caret)
#library(stepPlr)
#library(e1071)
#library(ranger)
#
# Set table theme to allow for plotmath expressions
#tt <- ttheme_default(colhead=list(fg_params = list(parse=TRUE)))
```

### San Francisco Crime Data

San Francisco crime data used for this example is publicly available at 
https://data.sfgov.org and can be
accessed via the [SODA Consumer API](https://dev.socrata.com/consumers/getting-started.html).
The `RSocrata` package is used here to interact with the API easily from R. The data set we
will be working with (crime data in the years 2003 through 2014) consists of about 1.7 million
events, so we will need to register for an API token and store it in the file
`../proprietary_data/DataSFAppToken.txt`. If you are not familiar with Socrata and the SODA
Consumer API, be sure to read the entire [getting started guide](https://dev.socrata.com/consumers/getting-started.html). 
(The link to create a developer
account and register an API token is at the bottom of the page.)

*NOTE: Depending on your internet connection speed, downloading the data in the next step  will take a while. Also, storing the data file will take about a half gigabyte of disk space.*

*NOTE: Working directory should be set to this file's location.*

```{r getData, cache=TRUE}
token <- scan("../proprietary_data/DataSFAppToken.txt",what = "character")
dataFile <- "../proprietary_data/socrata_crime.csv"

if(!exists("socrataDF")){ #check to see if dataframe exists in interactive environment
  if(file.exists(dataFile)){
    socrataDF <- read.csv(dataFile)
  } else { 
    socrataDF <- read.socrata("https://data.sfgov.org/resource/cuks-n6tp.csv?$where=date between '2003-01-01T00:00:00' and '2014-12-31T23:59:59'", app_token = token)
    write.csv(socrataDF, file = "../proprietary_data/socrata_crime.csv")
  }
}

socrataDF$datechar <- format(socrataDF$date,"%m/%d/%Y")
socrataDF$datetime <- as.POSIXct(strptime(paste(socrataDF$datechar, socrataDF$time), "%m/%d/%Y %H:%M"))
```

### Table 7.3 San Francisco crime data in its raw form, as a sequence of events

```{r table7.3, cache=TRUE, dependson="getData"}
tableRows <- 10
kable(head(
  socrataDF %>% 
    select(c(incidntnum, datetime, datechar, time, pddistrict, category)) %>%
    filter(datetime > as.POSIXct("2008-04-13") & datetime < as.POSIXct("2008-04-14")) %>%
    arrange(time) %>%
    filter(time > "00:50") %>%
    select(-datetime),
  tableRows), col.names = c("Incident number", "Date", "Time", "District", "Category"))
  
```

### Listing 7.4 Converting SF crime event data to classical time series

```{r listing7.4, cache=TRUE, dependson="getData"}
socrataDF$month <- floor_date(socrataDF$datetime, "month")
crimeByMonth <- socrataDF %>% group_by(month) %>% summarise(n = n())
crimeByMonth <- crimeByMonth[-c(1,146),] # remove "2002-12-01" data that is for a partial month and "NA"
crimeTS <- ts(crimeByMonth$n, frequency = 12, start = c(2003,1))
```

### Figure 7.6 Classical time series of monthly crime count in San Francisco.
```{r figure7.6, cache=TRUE, dependson="listing7.4"}
plot(crimeTS,
     xlab = "Year",
     ylab = "Number of crimes",
     axes=FALSE)
axis(side=1, at=c(2003:2015))
axis(side=2, at=seq(8000, 14000, by=1000))
box()
```

### Listing 7.5 Windowed statistics and differences

```{r listing7.5, cache=TRUE, dependson=listing7.4}
crimeWindow1 <- window(crimeTS,c(2014,3),c(2014,6)) #crime data from Spring, 2014
meanWindow1 <- mean(crimeWindow1)
stdWindow1 <- sd(crimeWindow1)
crimeWindow2 <- window(crimeTS,c(2013,3),c(2013,6)) #crime data from Spring, 2013
meanWindow2 <- mean(crimeWindow2)
stdWindow2 <- sd(crimeWindow2)
meanDiff <- meanWindow1 - meanWindow2
stdDiff <- stdWindow1 - stdWindow2

kable(data.frame(variable = c("meanWindow1", "meanWindow2", "meanDiff", "stdWindow1", "stdWindow2", "stdDiff"),
                 value = c(meanWindow1, meanWindow2, meanDiff, stdWindow1, stdWindow2, stdDiff)))

```

### Figure 7.7 Top: Correlation of the original time series and 12-month lagged time series defines the 12-month autocorrelation.  Bottom: The autocorrelation function for the SF crime data.

```{r figure7.7, cache=TRUE, dependson="listing7.4"}
crimeTSlagged <- lag(crimeTS, 12)
plot(crimeTS,
     xlab = "Year",
     ylab = "Number of crimes",
     axes=FALSE)
axis(side=1, at=c(2003:2015))
axis(side=2, at=seq(8000, 14000, by=1000))
box()
lines(crimeTSlagged, col = "blue")
#TODO add legend

acf(crimeTS, lag.max = 50)
```

### Figure 7.8 Left: Periodogram of the San Francisco crime data, showing the spectral density as a function of frequency. Right: The same periodogram with the x-axis transformed from frequency to period.

```{r figure7.8, cache=TRUE, dependson="figure7.7"}

spec.pgram(crimeTS, detrend = TRUE)

```



