---
title: 'Real-World Machine Learning (with R): Chapter 5'
author: "Paul Adamson"
date: "November 11, 2016"
output: html_document
---

This notebook contains R code to accompany Chapter 5 of the book 
["Real-World Machine Learning"](https://www.manning.com/books/real-world-machine-learning),
by  Henrik Brink, Joseph W. Richards, and Mark Fetherolf.  It is part of a 
series of R Markdown notebooks hosted on GitHub in the 
[rwml-R repo](https://github.com/padamson/rwml-R).

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(bit64)
library(dplyr)
library(caret)
workDir <- "/Users/padamson/analytics/real-world-machine-learning-R/"
```

### Data

The data for the examples in this chapter are from the
[Kaggle Event Recommendation Engine Challenge](https://www.kaggle.com/c/event-recommendation-engine-challenge).
The rules of the challenge prohibit redristribution of the data. 
Login to [Kaggle](https://www.kaggle.com), download the data, and put
the `train.csv`, `events.csv`, and `users.csv` files in a folder 
named `./proprietary_data/`.

### Figure 5.2 A sample of the datasets used for training the event-recommendations model

```{r loadData, cache=TRUE, dependson="setup"}
dataDir <- "proprietary_data"
# join on user and event identifiers
#from users data, get user_id, birthyear, gender, timezone
users <- read.csv(file.path(dataDir, "users.csv"),
                  colClasses = c('integer64', 
                                 "NULL", 
                                 "integer",
                                 "factor",
                                 "NULL",
                                 "NULL",
                                 "integer"),
                  na.strings = c("16-Mar", "23-May", "NA", "None"))

# from events data, get event_id, lat, lng
# note don't get confused about the user_id in this dataset. it is the user that created the event and is different from the user_id in the train dataset
events <- fread(file.path(dataDir, "events.csv"),
                sep=",",
                colClasses = c("integer64",
                               rep("NULL",6),
                               "numeric",
                               "numeric",
                               rep("NULL",101)))

str(events)

# from train data, get user, event, invited, interested, not_interested 
# rename user to user_id and event to event_id to match users and events datasets
# also need to drop observations without explicit interested or not_interested selection
# then drop not_interested since we only need one of the categorical variables
train <- read.csv(file.path(dataDir, "train.csv"),
                  colClasses = c("integer64",
                                 "integer64",
                                 "integer",
                                 "NULL",
                                 "integer",
                                 "integer"),
                  col.names = c("user_id",
                                "event_id",
                                "invited",
                                "NULL",
                                "interested",
                                "not_interested")) %>%
  subset(interested == 1 | not_interested == 1) %>%
  select(-not_interested)
```

```{r figure5.2, cache=TRUE, dependson="loadData"}
train.joined <- inner_join(train, users, by=c("user_id")) %>%
  inner_join(events, by=c("event_id")) %>%
  select(-c(user_id,event_id))
kable(head(train.joined,7))
```

## Figure 5.3 Cross-validated ROC curve and AUC metric for the simple event- recommendation model
```{r figure5.3, cache=TRUE, dependson="figure5.2"}
dummies <- dummyVars(" ~ .", data = train.joined, fullRank = TRUE, 
                     levelsOnly = TRUE)
train.joined.numeric <- data.frame(predict(dummies, newdata = train.joined))
kable(head(train.joined.numeric,7))
```
